---
title: "*mineMS2*: Mining of MS fragmentation spectra by frequent subgraph mining."
author: "Alexis Delabriere and Eteinne Thevenot"
date: "`r Sys.Date()`"

vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: "mineMS2-vignette.bib"
output: 
  BiocStyle::html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=6, fig.path='figures/')
```

# Context

	
# Initialization
We first load the MineMS2 package
```{r mineMS2_loading}
library(mineMS2)
```

##Dataset
A dataset of demonstration is available in this package, it contains 

# Input
The _mineMS2_ package take as input any of the following format `r recognisedFormat()` and mays also take as input a list _Spectrum2_ object from the _MSnbase_ package. The initial spectra reading is done using the *ms2Lib* constructor.
function.
```{r paths}
# Finding the data
path_demo <- system.file("dataset",package="mineMS2")
ex_file <- file.path(path_demo,"ex_mgf.mgf")
```
The spectra information are centralized in an *ms2Lib* object. It can take 3 kind of inputs :

* A vector of filename with compatible format (`mineMS2:::recognisedFormat()`)
	
* A single directory name in which the files in supported format are stored.
	
* A list containg Spectrum2 objects.
	
This object store all the supplementary information
	

```{r constructor_ms2lib}
m2l <- ms2Lib(ex_file)
m2l
```
At this step some supplementary informations may be added to the spectra using the _spectraInfo_ arguments. It should be a table with as much row as spectra, in the same order.
```{r constructor_ms2lib_supp_info}
###Loading the suppelmentary information
supp_infos_path <- file.path(path_demo,"ex_supp_infos.csv")
supp_infos <- read.table(supp_infos_path,header=TRUE,sep=";")
m2l <- ms2Lib(ex_file,suppInfos = supp_infos)
m2l
```

These supplementary information may also be used to generate an ID for each spectra, which can be used in the functions of the MineMS2 software, and which is reused into the visualisation. Here fore examples we design an ID composed of the mass and the retention time for the precursor for each spectrum. Initially this ID is set to S1,S2,...,SN where N is the number of file by the mineMS2 software.
```{r spectrum_id}
infos <- getInfo(m2l,"S")
ids <- paste(paste("MZ",infos[,"mz"],sep=""),paste("RT",infos[,"rt"],sep=""),sep="_")
m2l <- setIds(m2l,ids)
```
This ID is notably used in the visualisation offered by the *mineMS2* package.


The extracted spectrum may be plot using the *plot* function of the *msnBase* package :
```{r plotSpec}
plot(m2l["S10"],full=TRUE)
```
Here the "S10" index correspond to the first spectra. These indexes will be discussed later.


#Losses dicretization
The graphs are built by discretizing the mass losses. This algorithm uses a density with a varying bandwidth to take into account the fact that the mass accuraccy of mass spectrometers are in ppm. This step is very important, as it determine the initial set of graphs. The most important parameters and their meaning are listed here :

* _ppm_: The accuracy of the mass spectrometer in ppm used to determine the bandwidth of the density.

* _dmz_: The minimum deviation use for bandwidth in dalton.

* _count_: The minimum number of spectra in which a loss needs to be present.

* _precPpm_: The ppm tolerance used to match the precursor.

* _precDmz_: The minimum tolerance used to match the precursor.

* _heteroAtoms_: Shall hetero atoms (*P*,*S*,*Cl*) be used when generating ofrmula for mass losses.

* _maxFrags_: Only the _maxFrags_ most intense will be ocnsidered to build the graph. At most 20.


The _precPpm_ and _precDmz_ arguments are used to find a peak corresponding to the precursor on the MS-MS spectra. If it is not present an artificial peak is added.

```{r edge_discretization,message=FALSE}
m2l <- discretizeMassLosses(m2l,ppm=8,dmz=0.008,count=2,
							precPpm=20,precDmz=0.03,maxFrags=15)
```
After this step, the discretized losses labels may be extracted :
```{r tab_losses}
head(lossesTable(m2l))
```
After this steps the graphs correspoding to the spectra have been constructed. 

It is good to note that at this step supplementary information may using the penalizedLosses informations.

#Mining frequent subgraphs 
To generate the frequent subgraph we will use the *mineClosedSubgraphs* function. Two important parameters are the _count_ parameter which is the minimum number of occurences of a patterns, and the _sizeMin_ parameter which determine the minimum size of the mined graphs. Here we set _sizeMin_ to 1 to mine single losses.

```{r graph_mining}
m2l <- mineClosedSubgraphs(m2l,count=2,sizeMin = 1)
```
This step can be quite time consuming.

#IDs
The *ms2Lib* object include 3 different data types an they can be queried using their IDs which include a prefix specifying the data type and a number. These data types are the spectra (S), the patterns (P) and the losses (L). Further information on the functionality of these package are given in the next subsection :

##Spectra
Spectra may be queried using id starting by "S", they may be visualised using the "plot" method of "Spectrum2" object form the "msnBase" package. It is possible to find a specific spectra by using the _findMz_ function, with the "S" parameter.
```{r spec_plot_example}
sid <- findMz(m2l,147.0,type="S",ppm=8,dmz=0.2)
sid
```
Informations about the spectra may be queried using the *getInfo* function :
```{r get_info_s}
getInfo(m2l,sid)
```
The patterns in which the psectra is present may be extracted using the the *select* function : 

```{r select_s_p}
pids <- select(m2l,sid,"P")
pids
```

##Losses

Losses IDs start by the *L* letter. The can be searched using the *findMz* function to. Some information may be extracted using the *getInfo* function. We don't demonstrate it as they are similar to those of spectra. They can be searched against patterns (*P*) or dags (*D*) using the *select* function :
```{r select_l_p}
pids <- select(m2l,"L5","P")
```

##Patterns

* Patterns IDs start by the *P* letter, they can be plotted using the *plot* and *plotOccurences* functions.
```{r patterns_plot_example}
plot(m2l,"P80")

```
```{r patterns_plot_occs_example}
plotOccurences(m2l,"P10",byPage=9)
```


# Session info

Here is the output of `sessionInfo()` on the system on which this document was
compiled:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
