---
title: 'MineMS2: Coupling with GNPS'
author: "Alexis Delabriere"
date: "22 octobre 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# Introduction

This vignette highlights the use of the _MineMS2_ software coupled to the GNPS MS-MS networking methodology. The GNPS network may be generated directly from the *ex_mgf.mgf* file present in the *dataset* subdirectory of your _MineMS2_ installation. In this example the GNPS network has been precomputed on the GNPS website and extracted, the *.graphML* file  *ex_gnps_network.graphml* is present inside the *dataset* subdirectory.


```{r path_gnps}
path_demo <- system.file("dataset",package="mineMS2")
ex_network <- file.path(path_demo,"ex_gnps_network.graphml")
```


#Generating mineMS2 patterns.

The patterns are generated using the full datasets, and are therefore not necessarely.

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
