---
title: "*mineMS2*: Mining of MS fragmentation spectra by frequent subgraph mining."
author: "Alexis Delabriere and Eteinne Thevenot"
date: "`r Sys.Date()`"

vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: "mineMS2-vignette.bib"
output: 
  BiocStyle::html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=6, fig.path='figures/')
```

# Context
Mining interest blablablabla
	* Grpha natural representation. 
	* Efficient algorithm
	* Blablabla
	
# Importing the files
```{r loading, echo=TRUE, warning=FALSE, message=FALSE}
# loading the package
library(mineMS2)
```
```{r paths}
# Findinf the data
path_demo <- system.file("mgf",package="mineMS2")
print(head(list.files(path_demo)))
```
The spectra information are centralized in an *ms2Lib* object. It can take 3 kind of inputs :

	* A vector of filename with compatible format (`mineMS2:::recognisedFormat()`)
	
	* A single directory name in which the files in supported format are stored.
	
	* A list containg Spectrum2 objects.
	

```{r constructor_ms2lib,fig.show="hide",message=FALSE}
m2l <- ms2Lib(path_demo)
m2l

```


#Building the graphs
The graph are built by discrtiezing the mass losses.
```{r edge_discrtization}

m2l <- discretizeMassLosses(m2l,8,0.008,2/length(m2l),precPpm=20,precDmz=0.03,maxFrags=15)

```


# Session info

Here is the output of `sessionInfo()` on the system on which this document was
compiled:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
